# Sampling frames



We will have different sampling frames, also called the _universe_, for the different ANO-modules.

## Setup schema

```{r}
schema_ur <- "CREATE SCHEMA sampling_frames"
dbSendQuery(con, schema_ur) 
```
Write queries to grant read only access to all.
```{r}
priv <- "ALTER DEFAULT PRIVILEGES IN SCHEMA sampling_frames GRANT SELECT ON TABLES TO ag_pgsql_ano_moduler_ro"
priv2 <- "ALTER DEFAULT PRIVILEGES IN SCHEMA sampling_frames GRANT SELECT ON TABLES TO ag_pgsql_ano_moduler_rw"
priv3 <- "ALTER DEFAULT PRIVILEGES IN SCHEMA sampling_frames GRANT SELECT ON TABLES TO ag_pgsql_ano_moduler_admin"
priv4 <- "GRANT USAGE ON SCHEMA sampling_frames  TO ag_pgsql_ano_moduler_admin"
priv5 <- "GRANT USAGE ON SCHEMA sampling_frames  TO ag_pgsql_ano_moduler_rw"
priv6 <- "GRANT USAGE ON SCHEMA sampling_frames  TO ag_pgsql_ano_moduler_ro"

dbSendStatement(con, priv)
dbSendStatement(con, priv2)
dbSendStatement(con, priv3)
dbSendStatement(con, priv4)
dbSendStatement(con, priv5)
dbSendStatement(con, priv6)
```



## ANO Havstrand

For ANO Havstrand (ANO Coast), SSB500 has been masked to only include grid cells that overlap with the Norwegian coastline.
The script for doing that is written in python ad can be found here: */data/P-Prosjekter2/412421_okologisk_tilstand_2024/Jan/*

Read data into R:

```{r}
#| eval: true
coast_path <- "/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Jan/CoastalSampling3.gdb"

sf::st_layers(coast_path)
```

SSB500_Coast4 contains the SSB500 grid cells that overlap with the coastline.

```{r}
coastSSB <- read_sf(coast_path, layer = "SSB500_Coast4")
```

The SSBid will be the foregn key, linking to ssb_500.
We also need a primary key, and therefore I add a unique identifier.

```{r}
ids <- UUIDgenerate(n = nrow(coastSSB))
anyDuplicated(ids)
coastSSB <- coastSSB |>
  select(ssbid = SSBid)|>
  mutate(havstrand_id = ids)

```



We don't need the geometry
```{r}
coastSSB <- as_tibble(coastSSB) |>
  select(ssbid, havstrand_id)

```

### Define table properties
We will name the table *samplingframe_havstrand_2024*.

```{r}
q1 <- "create table sampling_frames.samplingframe_havstrand_2024 (
havstrand_id character varying(50) primary key,
ssbid character varying(50),
CONSTRAINT fk_ssb_500
      FOREIGN KEY (ssbid)
        REFERENCES ssb_grids.ssb_500 (ssbid)
);"

# indices makes the database work faster. It should be added to all tables that are looked up frequently
q2 <- "create index on sampling_frames.samplingframe_havstrand_2024 using btree(ssbid);"
q3 <- "create index on sampling_frames.samplingframe_havstrand_2024 using btree(havstrand_id);"
```

```{r}
# sending the queries:
dbSendStatement(con, q1)
dbSendStatement(con, q2)
dbSendStatement(con, q3)
```


### Write to db

Then we write data to the samplingframe_havstrand_2024 table.
```{r}
write_sf(coastSSB, dsn = con,
         layer = Id(schema = "sampling_frames", table = "samplingframe_havstrand_2024"), 
         append = T)
```